pipeComp, a general framework for the evaluation of computational pipelines, reveals performant single cell RNA-seq preprocessing tools

### 2020-05-13 bmc_article.tex
Updates from Overleaf

\begin{abstract} % abstract
%\parttitle{Background} %if any
%The massive growth of single-cell RNA-sequencing (scRNAseq) and methods for its analysis still lacks sufficient and up-to-date benchmarks that would guide analytical choices. Moreover, current studies are often focused on isolated steps of the process. Here, we present a flexible R framework for pipeline comparison with multi-level evaluation metrics and apply it to the benchmark of scRNAseq analysis pipelines using datasets with known cell identities. We evaluate common steps of such analyses, including filtering, doublet detection, normalization, feature selection, denoising, dimensionality reduction and clustering. On the basis of these analyses, we make a number of concrete recommendations about current analysis choices. The evaluation framework, \textit{pipeComp}, has been implemented so as to easily integrate any other step or tool, allowing extensible benchmarks and easy application to other fields (\url{https://github.com/plger/pipeComp}).

@@

{\color{red}We next investigated the impact of several filtering methods (see Methods for more details): four methods based on deviations to MADs with increasing levels of stringency (named \textit{lenient}, \textit{default}, \textit{stringent}, \textit{veryStringent}) and two methods based on \textit{scater}'s \textit{runPCA} using all (\textit{pca.all}) or selected covariates (\textit{pca.sel})}. An important risk of excluding cells on the basis of their distance from the whole distribution of cells on some properties (e.g., library size) is that these properties tend to have different distributions across subpopulations. As a result, thresholds in terms of number of MADs from the whole distribution can lead to strong biases against certain subpopulations (Figure \ref{fig:filtering}A). We therefore examined the tradeoff between the increased accuracy of filtering and the maximum proportion of cells excluded per subpopulation (Figure \ref{fig:filtering}B and Supplementary Figure 11). Since filtering changes the relative abundance of the different subpopulations, global clustering accuracy metrics such as ARI are not appropriate here. We therefore calculated the per-subpopulation precision and recall using the Hungarian algorithm \cite{PapadimitriouHu1998} and monitored the mean F1 score {\color{red}across subpopulations. A first observation was that, although more stringent filtering tended to be associated with an increase in accuracy, it tended to plateau and could also become deleterious. Most of the benefits could be achieved without very stringent filtering and minimizing subpopulation bias (Figure \ref{fig:filtering}B). Applying the same filtering criteria on individual clusters of cells (identified through scran's \texttt{quickCluster} method) rather than on whole dataset resulted in nearly no cells being filtered out (Supplementary Figure 12A). This suggests that stringent filtering on the global population tends to discard cells of subpopulations with more extreme properties (e.g. high library size), rather than low-quality cells. In contrast, too lenient filtering (or lack thereof) can lead to misclassified cells, as was the case for the Zhengmix4 datasets (Figure \ref{fig:filtering}B). We found that a relatively mild filtering (`default' - see methods) provided a good trade-off, often leading to high clustering performance (e.g. simMix and Zhengmix4 datasets) while retaining most of the cells in each subpopulation (e.g. Koh and Zhengmix8eq). PCA-based selection did not appear superior to feature-wise filters. We found very little overlap between the cells excluded by doublet removal and those excluded by MAD-based filtering (Supplementary Figure 12B). Of note, doublet removal in conjunction with filtering tended to improve accuracy not only in datasets that do include heterotypic doublets (mixology datasets), but also some others in which such doublets are unlikely given the experimental design (e.g. FACS-sorted datasets). Indeed, the datasets where doublet removal did not have a clear positive impact (e.g. Koh, Kumar, and the simulations) are those in which heterotypic doublets are not expected. Overall, we therefore recommend the use of doublet removal followed by our `default' filtering (see Methods) or a similar approach.}

@@

{\color{red}\subsection*{Application of \textif{pipeComp} to a different context}

@@

\begin{enumerate}
   \item Filtering
   \begin{itemize}
     \item Doublet detection and removal is advised and can be performed at little computing cost with software such as \textit{scDblFinder} or \textit{scds}.
     \item Distribution-based cell filtering fails to capture doublets and should use relatively lenient cutoffs (e.g., 5 MADs, or 3 MADs in at least 2 distributions \textcolor{ref}{- see for instance our 'default' filters in the methods}) to exclude poor-quality cells while avoiding bias against some subpopulations.
     \item Features filtering based on feature type did not appear beneficial.
   \end{itemize}
   \item Normalization and scaling
   \begin{itemize}
     \item Most normalization methods tested yielded a fair performance, especially when combined with scaling, which tended to have a positive impact on clustering.
     \item \textit{SCTransform} offered the best overall performance in terms of the separability of the subpopulations, as well as removing the effect of library size and detection rate.
     \item The common practice of regressing out cell covariates, such as the detection rate or proportion of mitochondrial reads nearly always had a negative impact, leading to increased correlation with covariates and decreased clustering accuracy. We therefore advise against this practice.
   \end{itemize}
   \item Feature selection
   \begin{itemize}
     \item Deviance \cite{townesGlmpca2019} offered the best ranking of genes for feature selection.
     \item Increasing the number of features included tended to lead to better classifications, plateauing from 4000 features in our datasets.
   \end{itemize}
   \item Denoising/imputation
   \begin{itemize}
       \item Denoising appeared beneficial to the identification of some subpopulations in 10x datasets, but not in Smart-seq datasets.
       \item We found especially \textit{ALRA} (with prior normalization), \textit{DrImpute} (with prior processing) and \textit{DCA} to offer the best performances\textcolor{red}{, although none of these methods was beneficial over all subpopulations}.
   \end{itemize}
   \item PCA
   \begin{itemize}
     \item Similarly to previous reports \cite{SunDimRed2019}, we recommend the irlba-based PCA using weighting of the components, as implemented in \textit{Seurat}. 
     \item We advise against the commonly-used elbow method (because it is too conservative) and the jackstraw method (low performance at great computational costs) for deciding on the components to include, and recommend a method like the \textit{global maximum likelihood based on translated Poisson Mixture Model} method (e.g., implemented in the  \texttt{maxLikGlobalDimEst} function of \texttt{intrinsicDimension}, in our case using the \textcolor{red}{10 or }20 nearest neighbors).
   \end{itemize}
   \item Clustering
   \begin{itemize}
    \item We found \textit{scran}-based \textit{walktrap} clustering to show good performances, at however increased computational costs which might be prohibitive with large datasets.
    \item In cases where prior knowledge can guide the choice of a resolution, \textit{Seurat} can be useful in affording manual control of it while, in the absence of such knowledge, \textit{scran}-based \textit{walktrap} clustering provided reasonable estimates.
   \end{itemize}
\end{enumerate}

>>>>

\begin{abstract}

@@

{\color{red}We next investigated the impact of several filtering methods (see Methods for more details): four methods based on deviations to MADs with increasing levels of stringency (named \textit{lenient}, \textit{default}, \textit{stringent}, \textit{veryStringent}) and two methods based on \textit{scater}'s \textit{runPCA} using all (\textit{pca.all}) or selected covariates (\textit{pca.sel})}. An important risk of excluding cells on the basis of their distance from the whole distribution of cells on some properties (e.g., library size) is that these properties tend to have different distributions across subpopulations. As a result, thresholds in terms of number of MADs from the whole distribution can lead to strong biases against certain subpopulations (Figure \ref{fig:filtering}A). We therefore examined the tradeoff between the increased accuracy of filtering and the maximum proportion of cells excluded per subpopulation (Figure \ref{fig:filtering}B and Supplementary Figure 11). Since filtering changes the relative abundance of the different subpopulations, global clustering accuracy metrics such as ARI are not appropriate here. We therefore calculated the per-subpopulation precision and recall using the Hungarian algorithm \cite{PapadimitriouHu1998} and monitored the mean F1 score {\color{red}across subpopulations. A first observation was that, although more stringent filtering tended to be associated with an increase in accuracy, it tended to plateau and could also become deleterious. Most of the benefits could be achieved without very stringent filtering and minimizing subpopulation bias (Figure \ref{fig:filtering}B). Applying the same filtering criteria on individual clusters of cells (identified through scran's \texttt{quickCluster} method) rather than on whole dataset resulted in nearly no cells being filtered out (Supplementary Figure 12A). This suggests that stringent filtering on the global population tends to discard cells of subpopulations with more extreme properties (e.g. high library size), rather than low-quality cells. In contrast, too lenient filtering (or lack thereof) can lead to misclassified cells, as was the case for the Zhengmix4 datasets (Figure \ref{fig:filtering}B). We found that a relatively mild filtering (\textit{default} - see methods) provided a good trade-off, often leading to high clustering performance (e.g. simMix and Zhengmix4 datasets) while retaining most of the cells in each subpopulation (e.g. Koh and Zhengmix8eq). PCA-based selection did not appear superior to feature-wise filters. We found very little overlap between the cells excluded by doublet removal and those excluded by MAD-based filtering (Supplementary Figure 12B). Of note, doublet removal in conjunction with filtering tended to improve accuracy not only in datasets that do include heterotypic doublets (mixology datasets), but also some others in which such doublets are unlikely given the experimental design (e.g. FACS-sorted datasets). Indeed, the datasets where doublet removal did not have a clear positive impact (e.g. Koh, Kumar, and the simulations) are those in which heterotypic doublets are not expected. Overall, we therefore recommend the use of doublet removal followed by our \textit{default} filtering (see Methods) or a similar approach.}

@@

{\color{red}\subsection*{Application of \textit{pipeComp} to a different context}

@@

\begin{enumerate}
   \item Filtering
   \begin{itemize}
     \item Doublet detection and removal is advised and can be performed at little computing cost with software such as \textit{scDblFinder} or \textit{scds}.
     \item Distribution-based cell filtering fails to capture doublets and should use relatively lenient cutoffs (e.g., 5 MADs, or 3 MADs in at least 2 distributions \textcolor{ref}{- see for instance our \textit{default} filters in the methods}) to exclude poor-quality cells while avoiding bias against some subpopulations.
     \item Features filtering based on feature type did not appear beneficial.
   \end{itemize}
   \item Normalization and scaling
   \begin{itemize}
     \item Most normalization methods tested yielded a fair performance, especially when combined with scaling, which tended to have a positive impact on clustering.
     \item \textit{SCTransform} offered the best overall performance in terms of the separability of the subpopulations, as well as removing the effect of library size and detection rate.
     \item The common practice of regressing out cell covariates, such as the detection rate or proportion of mitochondrial reads nearly always had a negative impact, leading to increased correlation with covariates and decreased clustering accuracy. We therefore advise against this practice.
   \end{itemize}
   \item Feature selection
   \begin{itemize}
     \item Deviance \cite{townesGlmpca2019} offered the best ranking of genes for feature selection.
     \item Increasing the number of features included tended to lead to better classifications, plateauing from 4000 features in our datasets.
   \end{itemize}
   \item Denoising/imputation
   \begin{itemize}
       \item Denoising appeared beneficial to the identification of some subpopulations in 10x datasets, but not in Smart-seq datasets.
       \item We found especially \textit{ALRA} (with prior normalization), \textit{DrImpute} (with prior processing) and \textit{DCA} to offer the best performances\textcolor{red}{, although none of these methods was beneficial over all subpopulations}.
   \end{itemize}
   \item PCA
   \begin{itemize}
     \item Similarly to previous reports \cite{SunDimRed2019}, we recommend the irlba-based PCA using weighting of the components, as implemented in \textit{Seurat}. 
     \item We advise against the commonly-used elbow method (because it is too conservative) and the jackstraw method (low performance at great computational costs) for deciding on the components to include, and recommend a method like the \textit{global maximum likelihood based on translated Poisson Mixture Model} method (e.g., implemented in the  \texttt{maxLikGlobalDimEst} function of \texttt{intrinsicDimension}, in our case using the \textcolor{red}{10 or }20 nearest neighbors).
   \end{itemize}
   \item Clustering
   \begin{itemize}
    \item We found \textit{scran}-based \textit{walktrap} clustering to show good performances, at however increased computational costs which might be prohibitive with large datasets.
    \item In cases where prior knowledge can guide the choice of a resolution, \textit{Seurat} can be useful in affording manual control of it while, in the absence of such knowledge, \textit{scran}-based \textit{walktrap} clustering provided reasonable estimates.
   \end{itemize}
\end{enumerate}
